Optimal (Multiway) Spatial Joins

### 2024-05-24 msj.tex
yufei

\section{H-V Multiway Spatial Joins} \label{sec:hv}

This section deals with a special version of the $k$-SJ problem. Recall that the input of $k$-SJ comprises $k$ sets of rectangles: $R_1, R_2, ..., R_k$. In the special version --- which we name the {\em H-V $k$-SJ} problem ---  we impose the constraints that (i) $k \ge 3$, (ii) $R_{k-1}$ should be a set of horizontal segments, and (iii) $R_k$ should be a set of vertical segments. For better clarity, we will represent the input sets as $R_1, R_2, ..., R_{k-2}, H$ ($=R_{k-1}$), and $V$ ($=R_k$). Remember that each $R_i$ ($i \in [k-2]$) is a set of rectangles. As shown in the next section, solving the H-V $k$-SJ problem is an imperative step for settling the original $k$-SJ problem.

@@

Our objective is to prove that the H-V $k$-SJ problem can be efficiently reduced to the $(k-1)$-SJ problem --- note: it is $(k-1)$-SJ here, rather than H-V $(k-1)$-SJ. For this purpose, we will assume the existence of an algorithm $\A$ that can solve $\kappa$-SJ problems for all $\kappa \in [2, k-1]$. Denote by $F_\kappa(n, \out)$ the worst-case runtime of $\A$ on any instance of $\kappa$-SJ that has input size $n$ and output size $\out$. We consider that $F_\kappa(n, \out) \le F_{\kappa + 1}(n, \out)$, that is, its overhead on $\kappa$-SJ should not be larger than that on $(\kappa+1)$-SJ.

@@

The rest of the section serves as a proof for:

Recall that $n$ is the total size of $R_1, ..., R_{k-2}, H$, and $V$. Henceforth, we assume that a distinct ID from 1 to $n$ is given to each rectangle or segment in $R_1, ..., R_{k-2}, H$, and $V$.

\subsection{Categorizing the Result Tuples} \label{sec:hv:result-part}

Let us represent the result of the H-V $k$-SJ problem as $\J(R_1, ..., R_{k-2},$ $H, V)$. Recall that $\J(R_1, ..., R_{k-2},$ $H, V)$ contains all such $k$-tuples $(r_1, ..., r_{k-2}, h, v) \in R_1 \times ... \times R_{k-2} \times H \times V$ satisfying the condition that $h \cap v \bigcap_{i=1}^{k-2} r_i$ is not empty. The condition is equivalent to saying that the intersection point of $h$ and $v$ falls in every $r_i$ of $i \in [k-2]$.

Let us focus on any $k$-tuple $(r_1, ..., r_{k-2}, h, v) \in  \J(R_1, ..., R_{k-2},$ $H, V)$. We classify the tuple into one of the two categories below:

    \item {{\bf Type 2:}} $h$ or $v$ fails to cross some rectangle in $\set{r_1, r_2, ..., r_{k-2}}$. This means the existence of at least one rectangle $r_i$ (for some $i \in [k-2]$) that covers an endpoint of either $h$ or $v$ or both.
}
Figure~\ref{fig:hv:types} illustrates a result tuple of each type, assuming $k = 4$.

We will find result tuples of each type separately. Before proceeding, however, let us observe a sense of symmetry for Type 2. As mentioned, for a result tuple $(r_1, ..., r_{k-2}, h, v)$ of this type, a rectangle $r_i$, for some $i \in [k-2]$, covers an endpoint of $h$ or $v$ or both. As (i) there are $k-2$ choices for $i$ and (ii) $h$ and $v$ together have four endpoints, we can divide Type 2 further into $4(k-2)$ ``sub-types'': e.g., in subtype 1 (resp., 2), $r_1$ covers the left (resp., right) endpoint of $h$, in subtype 3 (resp., 4), $r_1$ covers the bottom (resp., top) endpoint of $v$, in subtype 5 (resp., 6), $r_2$ covers the left (resp., right) endpoint of $h$, etc. It is possible for the result tuple to belong to multiple sub-types simultaneously.

@@

\begin{figure} 
    \begin{tabular}{cc} 
        \includegraphics[height=27mm]{./artwork/type1} & 
        \hspace{5mm} \includegraphics[height=27mm]{./artwork/type2} \\
        (a) Type 1 result tuple &
        (b) Type 2 result tuple
    \end{tabular}

    \figcapup 
    \caption{Two categories of H-V $k$-SJ result tuples ($k$ = 4)} 
    \label{fig:hv:types}
    \figcapdown
\end{figure}

>>>>

\section{The Core: H-V Multiway Spatial Joins} \label{sec:hv}

Recall that the input of $k$-SJ comprises $k$ sets of rectangles: $R_1, R_2, ...,$ $R_k$. We now formulate a special version of $k$-SJ, named the {\em H-V $k$-SJ} problem. The special nature is reflected in the introduction of three constraints: (i) $k \ge 3$, (ii) $R_{k-1}$ should be a set of horizontal segments, and (iii) $R_k$ should be a set of vertical segments. For better clarity, we will represent the input sets as $R_1, R_2, ..., R_{k-2}, H$ ($=R_{k-1}$), and $V$ ($=R_k$). We want to output the join result $\J(R_1, ..., R_{k-2},$ $H, V)$, including every $k$-tuple $(r_1, ..., r_{k-2}, h, v) \in R_1 \times ... \times R_{k-2} \times H \times V$ such that $h \cap v \cap \bigcap_{i=1}^{k-2} r_i$ is not empty. As we will see later, conquering this sepcial version is the most challenging step in settling the original $k$-SJ problem optimally.

@@

Our objective is to prove H-V $k$-SJ can be efficiently reduced to $(k-1)$-SJ --- note: it is $(k-1)$-SJ here, rather than H-V $(k-1)$-SJ. For this purpose, we assume the existence of an algorithm $\A$ that can solve $\kappa$-SJ problems for all $\kappa \in [2, k-1]$. Denote by $F_\kappa(n, \out)$ the worst-case runtime of $\A$ on any instance of $\kappa$-SJ that has input size $n$ and output size $\out$. We consider that $F_\kappa(n, \out) \le F_{\kappa + 1}(n, \out)$, that is, its overhead on $\kappa$-SJ should not be larger than that on $(\kappa+1)$-SJ.
We will prove:

%The condition is equivalent to saying that the intersection point of $h$ and $v$ falls in every $r_i$ of $i \in [k-2]$.

To get started, consider any $k$-tuple $(r_1, ..., r_{k-2}, h, v)$ in the join result $\J(R_1, ..., R_{k-2}, H, V)$. We classify the tuple into one of the two types below:

    \item {{\bf Type 2:}} $h$ or $v$ fails to cross some rectangle in $\set{r_1, r_2, ..., r_{k-2}}$. Equivalently,
    there exists at least one rectangle $r_i$ (for some $i \in [k-2]$) that covers an endpoint of either $h$ or $v$ or both.
}
Figure~\ref{fig:hv:types} illustrates a result tuple of each type, assuming $k = 4$.

@@

We will find result tuples of each type separately. Before proceeding, however, let us observe a sense of symmetry for Type 2. As mentioned, for each result tuple $(r_1, ..., r_{k-2}, h, v)$ of this type, a rectangle $r_i$, for some $i \in [k-2]$, covers an endpoint of $h$ or $v$ or both. As (i) there are $k-2$ choices for $i$ and (ii) $h$ and $v$ together have four endpoints, we can divide Type 2 further into $4(k-2)$ ``sub-types'': e.g., in subtype 1 (resp., 2), $r_1$ covers the left (resp., right) endpoint of $h$, in subtype 3 (resp., 4), $r_1$ covers the bottom (resp., top) endpoint of $v$, in subtype 5 (resp., 6), $r_2$ covers the left (resp., right) endpoint of $h$, etc. It is possible for the result tuple to belong to multiple sub-types simultaneously.

@@

\begin{figure} 
    \begin{tabular}{cc} 
        \includegraphics[height=27mm]{./artwork/type1} & 
        \hspace{5mm} \includegraphics[height=27mm]{./artwork/type2} \\
        (a) Type 1 &
        (b) Type 2
    \end{tabular}

    \figcapup 
    \caption{Classifying H-V $k$-SJ result tuples ($k$ = 4)}
    \label{fig:hv:types}
    \figcapdown
\end{figure}
